Flooding of the Okavango Delta influences Connectivity for Dispersing African Wild Dogs

### 2023-01-05 Chapter_8/04_Manuscript/Manuscript.tex
various updates

During maximum flood, we observe a reduction in connectivity and an increase in
dispersal durations between distinct habitat patches. Notably, dispersal into
the center of the Okavango delta is reduced by 80\% during maximum flooding. At
minimum flooding, conversely, the delta reveals vital dispersal corridors and
increases chances of successful dispersal into neighboring habitat patches by xx
percent. Concomitantly, dispersal durations to move between patches is reduced,
on average by xx days,at low flood levels.

Climate change is expected to critically impact the flood-pulsing cycles of the
Okavango delta and may therefore substantially alter landscape connectivity in
this area. Our analysis suggests that anticipating such alterations in
environmental conditions is key when implementing conservation strategies.
Besides a better understanding of the conservation needs for the African wild
dog, our study also presents a first step towards incorporating environmental
changes due to seasonality or climate change in dispersal and connectivity
analyses and demonstrates the usefulness of individual-based dispersal
simulations as a pertinent conservation tool to study impacts of environmental
change on dispersal success.

@@

\section{Introduction}
Climate change is expected to profoundly impact ecosystems worldwide with
far-reaching consequences for species living therein \citep{Ozgul.2010,
Radchuk.2019, IPCC.2022}. By altering environmental conditions, climate change
affects animal behavior \citep{Fuller.2016}, resource availability
\citep{Durant.2007}, population dynamics \citep{Paniw.2021}, and the
distribution of wild living animal populations \citep{Thomas.2004,
Thuiller.2006}. One potential life-history pathway through which species may
mediate the effects of environmental change is dispersal \citep{Anderson.2012}.
By bringing individuals away from their natal location \citep{Clobert.2012},
dispersal equips species with a means to adapt to climate change by tracking
favorable habitat conditions \citep{Raia.2012} and to shift into a different
part of their fundamental niches \citep{Kokko.2022}. Dispersal furthermore
provides resilience against changing conditions by ensuring genetic diversity
\citep{Frankham.2002, Leigh.2012, Baguette.2013}, facilitating the colonization
empty habitats \citep{Gustafson.1996, Hanski.1999b, MacArthur.2001}, and
promoting the reinforcement of weakened and small subpopulations
\citep{Brown.1977}. As such, it is little surpsing that improving connectivity
is viewed as the number one strategy to combat the adverse effects of climate
change \citep{Heller.2009}.

While dispersal offers a means to offset the demographic consequences of climate
change \citep{Kokko.2006, Hodgson.2009, Travis.2013}, it itself is a function of
climatic and environmental conditions (e.g. \citep{Behr.2020}). The link between
dispersal and climaticor environmental can either be indirect, for example if
the propensity of individuals to disperse depends on environmental conditions,
or direct, when the biophysical environment through which dispersers move
affects dispersal prospects \citep{Travis.2013}. As the latter case indicates,
dispersal is also inextricably linked to the concept landscape connectivity
\citep{Baguette.2013}, which is understood as the degree by which the landscape
facilitates or impedes movements \citep{Taylor.1993}. Only if a sufficient
degree of landscape connectivity is maintained, successful dispersal of
individuals can be guaranteed \citep{Fahrig.2003}. Unfortunately, our
understanding how changing environmental conditions influence connectivity and
dispersal is limited \citep{Kokko.2006, Travis.2013}. Because collecting
information on dispersing animals is difficult and because it is difficult to
assess how changing environmental conditions affect connectivity...

The continued degradation and destruction habitats across the globe impedes
dispersal, promotes inbreeding and genetic drift and increases the
likelihood of local extirpation due to stochastic events \citep{Sawyer.2011}.
Combined, these factors limit species ability to adapt to changing environmental
conditions \citep{Fahrig.2003}.

@@

More recently, individual-based movement models (IBMMs) have been uprising and
are recommended for their ability to incorporate several crucial realities of
animal behavior, such as, for example, autocorrelation, that are by omitted by
necessity using resistance-based approaches \citep{Hofmann.2023,
Unnithan.2022a}. IBMMs allow much greater flexibility and are thus considered
powerful tools for examining connectivity under different landscape
configurations \citep{Littlefield.2019}.

Botswana is among the countries most heavily influenced by climate change .

@@

Talk about connectivity models
Talk about functional connectivity
Talk about habitat fragmentation and destruction
Say that wild dogs are keystone predators and an umberella species for
conservation

@@

The Okavango delta in Southern Africa poses a unique opportunity to study the
impacts of environmental change on species dispersal ability and connectivity in a
large scale natural experiment setup. According to projections by the IPCC,
Botswana and its surroundings are among the most vulnerable to climate change.
While global temperatures are expected to increase by xx degrees until the end
of the 21st century, For example, \cite{Engelbrecht.2015} predicts an increase
between 4 and 6\degree C for regions in southern Africa. One species that
heavily relies on dispersal for long-term persistence and that inhabits such an
arid ecosystem in Southern Africa is the African wild dog (\textit{Lycaon
pictus}). The African wild dog is a lightly built canid lives in cohesive packs
comprising up to 20 individuals. Wild dogs are social breeders and the majority
of reproduction is monopolized by the dominant alpha couple who are supported by
subordinates in taking care of their pups. The species exhibits strong
inbreeding avoidance, hence juvenile individuals born into the pack disperse in
same-sex sibling coalitions after reaching sexual maturity \citep{McNutt.1996,
Behr.2020}. Notably, the African wild dog is the sole extant representative of
its genus and thus considered as high-priority species for conservation efforts
\citep{Leigh.2012}. However, fragmentation and destruction of their historic
habitat may have decreased dispersal success for this species and thus
jeopardizes connectivity \citep{Leigh.2012}. Since the impacts of climate change
are not uniformly distributed across the planet, but particularly pronounced at
high latitudes and arid ecosystems, the species may be seen as most vulnerable
to such ecosystem shifts.

@@

In a previous study, \citep{Hofmann.2022} have investigated landscape
connectivity in the same ecosystem, yet assuming a static landscape.

Here, we capitalize on a previously parameterized dispersal model and use
dispersal simulations to investigate dispersal success and connectivity patterns
for African wild dogs under two extreme scenarios; one assuming maximum flooding
of the Okavango delta and one assuming minimum flooding of the Okavango delta.
Given that water poses a barrier to dispersing individuals, we anticipated that
dispersal prospects and connectivity during maximum flood are low. During
minimum flood, in contrast, we expect to reveal the presence of several
dispersal corridors that may be used to freely and quickly move between
remaining habitat patches.

\section{Materials and Methods}
We conducted all analyses using the programming language \textsf{R}. Any spatial
data manipulation was completed using the \textsf{terra} and \textsf{spatstat}
packages. Several helper functions for the simulation algorithm were written in
\textsf{C++} and imported to R using the \textsf{Rcpp} package. Figures were
generated using \textsf{ggplot2}, \textsf{igraph}, and \textsf{ggnetwork}. All
R-scripts used to conduct our analyses are provided through an online
repository.

@@

>>>>

During maximum flood, we observe a reduction in connectivity and an increase in
dispersal durations between distinct habitat patches. Notably, dispersal into
the central habitats of the Okavango delta is reduced by 80\% during maximum
flooding. At minimum flooding, conversely, the delta reveals vital dispersal
corridors and increases chances of successful dispersal into neighboring habitat
patches by xx percent. Furthermore, we find that dispersers get pushed into
different zones of potential human-wildlife conflict depending on the flood
extent.

Climate change is expected to critically impact water-cycles across the globe,
including the flood-pulsing dynamics of the Okavango delta. Our results suggest
that anticipating and taking into account such changes in environmental
conditions is key when developing and implementing conservation strategies,
especially when the aim is to improve connectivity. Besides this, our analysis
also provides important insights towards a better understanding of the
conservation needs for the African wild dog and demonstrates the usefulness of
individual-based dispersal simulations as a pertinent conservation tool to study
impacts of environmental change on dispersal success.

@@

\section{Introduction}
Climate change is expected to profoundly impact ecosystems worldwide with
far-reaching consequences for species living therein \citep{Ozgul.2010,
Radchuk.2019, IPCC.2022}. By altering environmental conditions, climate change
affects animal behavior \citep{Fuller.2016}, resource availability
\citep{Durant.2007}, population dynamics \citep{Paniw.2021}, and the
distribution of wild living animal populations \citep{Thomas.2004,
Thuiller.2006}. One life-history pathway through which species potentially
mediate the effects of environmental change is dispersal \citep{Anderson.2012},
i.e. the movement of individuals away from their natal location to the site of
first reproduction \citep{Clobert.2012}. The ability to disperse enables species
to adapt to climate change by tracking favorable habitat conditions
\citep{Raia.2012} or by shifting into a different region of their fundamental
niches \citep{Kokko.2022}. Dispersal furthermore provides resilience against
changing conditions by safeguarding genetic diversity \citep{Frankham.2002,
Leigh.2012, Baguette.2013}, facilitating the colonization empty habitats
\citep{Gustafson.1996, Hanski.1999b, MacArthur.2001}, and promoting the
reinforcement of weakened and small subpopulations \citep{Brown.1977}.

While dispersal offers a means to offset the demographic consequences of climate
change \citep{Kokko.2006, Hodgson.2009, Travis.2013}, it itself is a function of
climatic and environmental conditions (e.g. \citep{Elliot.2014, Behr.2020}). The
link between dispersal and the environment can either be indirect, for example
if the propensity of individuals to disperse depends on environmental
conditions, or direct, when the biophysical environment through which dispersers
move affects dispersal prospects \citep{Travis.2013}. The latter case highlights
that dispersal is also inextricably linked to the concept landscape connectivity
\citep{Baguette.2013}, which is understood as the degree by which the landscape
facilitates or impedes movements \citep{Taylor.1993}. A sufficient degree of
landscape connectivity is therefore a critical prerequisited of successful
dispersal \citep{Fahrig.2003}. Unfortunately, the continued degradation and
destruction habitats worldwide continues to imperil the dispersal ability of
many species and increases the prospects of inbreeding, genetic risk and of
local extirpation due to stochastic events \citep{Melbourne.2008, Sawyer.2011}.
Combined, these factors limit species ability to adapt to changing environmental
conditions \citep{Fahrig.2003}. Conservation strategies that aim to facilitate
dispersal by improving landscape connectivity are thus often viewed as pinacle
of conservation strategies to combat the adverse effects of climate change
\citep{Heller.2009}.

In recent years, there has been a growing body of research that uses animal
relocation data to identify movement corridors and assess connectivity at large
scales \citep{Chetkiewicz.2006, Doerr.2011, Squires.2013, Elliot.2014,
Benz.2016, Osipova.2019}. Identification of potential corridors relies on the
estimation of permeability (respectively resistance) surfaces, which return the
ease or willingness at which the focal species traverses a specific landscape
\citep{Sawyer.2011}. Such surfaces are typically created on the basis of habitat
preferences, which can be estimated using a suite of selection functions
\citep{Boyce.2002, Fortin.2005, Cushman.2010}. In selection functions, habitat
preferences are inferred by comparing spatial covariates (e.g. environmental and
anthropogenic) at locations visited by the animal to the same spatial covariates
at randomly selected locations in the animal's availability domain
\citep{Zeller.2012, Thurfjell.2014}. Importantly, selection functions rely on
adequate landscape and GPS relocation data that are representative of the
process being studied \citep{Diniz.2020}. More specifically, GPS relocation data
collected on dispersing individuals has been shown to outperform GPS relocation
data collected on resident individuals when aiming at detecting large scale
dispersal corridors \citep{Elliot.2014, Diniz.2020}. However, few connectivity
studies have partitioned their GPS relocation data according to behavioral modes
(e.g. resident vs. dispersing) of the studied species \citep{Wilson.2012,
Vasudev.2015}. As such, most permeability surfaces upon which dispersal
corridors are identified are created using GPS relocation data that were
collected on resident individuals. This introduces severe biases and
substantially reduces the power to reveal meaningful movement corridors, for
dispersing individuals have different needs and drives compared to resident
individuals \citep{Killeen.2014, Elliot.2014, Cozzi.2020}. Consequently, these
biases have limited our ability to assess the adequacy of the boundaries of
protected areas in securing connectivity for their protected species.

To study connectivity, various modeling techniques have emerged
\cite{Etherington.2016, Diniz.2019}. Initially, the techniques were limited to
studying strucutral aspects of connectivity by focusing on the composition and
configuration of habitat patches, while ignoring species' responses to the
landscape matrix \citep{Tischendorf.2000, Doerr.2011}. With the increasing
availability of telemetry data and methods to study habitat and movement
preferences \citep{Boyce.2002, Fortin.2009, Cushmann.2010, Avgar.2016}, however,
the focus has shifted towards more functional aspects of connectivity
\citep{Tischendorf.2000, Doerr.2011}. Currently, the most prominent
\textit{functional} connectivity models are least-cost path analysis (LCPA,
\citealp{Adriaensen.2003}) and circuit theory (CT, \citealp{McRae.2008}), two
graph-based methods that estimate conductance of the landscape by means of a
resistance (or inversely permeability) surface \citep{Zeller.2012}. Because both
approaches make assumptions that are rarely met by dispersing individuals,
individual-based movement models (IBMMs) have also gained some momentum recently
\citep{Kanagaraj.2013, Allen.2016, Hauenstein.2019, Diniz.2019, Zeller.2020,
Unnithan.2021, Unnithan.2022, Hofmann.2022}. In these models, movement of
individuals is explicitly simulated and connectivity metrics are derived based
on the simulated dispersal paths. IBMMs provide great modeling flexibility and
are thus considered powerful tools for examining connectivity under different
landscape configurations \citep{Littlefield.2019}.

Irrespective of the chosen method, connectivity is often modeled as a snapshot
in time that fails to render dynamic environmental change. However, some recent
studies have shown that connectivity may differ substantially depending on
season ... some examples.

 Unfortunately, our
understanding how changing environmental conditions influence connectivity and
dispersal is limited \citep{Kokko.2006, Travis.2013}. Because collecting
information on dispersing animals is difficult and because it is difficult to
assess how changing environmental conditions affect connectivity...

@@

More recently, individual-based movement models (IBMMs) have been uprising and
are recommended for their ability to incorporate several crucial realities of
animal behavior, such as, for example, autocorrelation, that are by omitted by
necessity using resistance-based approaches \citep{Hofmann.2023,
Unnithan.2022a}.
Botswana is among the countries most heavily influenced by climate change .

@@

Say that wild dogs are keystone predators and an umberella species for
conservation

@@

The Okavango delta in Southern Africa poses a unique opportunity to study the
impacts of environmental change on species dispersal ability and connectivity in
a large scale natural experiment setup. According to projections by the IPCC,
Botswana and its surroundings are among the most vulnerable to climate change.
While global temperatures are expected to increase by xx degrees until the end
of the 21st century, For example, \cite{Engelbrecht.2015} predicts an increase
between 4 and 6\degree C for regions in southern Africa. One of the keyspone
predators and an umberella species for conservation efforts in this ecosystem is
the African wild dog (\textit{Lycaon pictus}). This species heavily relies on
dispersal for long-term persistence. The African wild dog is a lightly built
canid lives in cohesive packs comprising up to 20 individuals. Wild dogs are
social breeders and the majority of reproduction is monopolized by the dominant
alpha couple who are supported by subordinates in taking care of their pups. The
species exhibits strong inbreeding avoidance, hence juvenile individuals born
into the pack disperse in same-sex sibling coalitions after reaching sexual
maturity \citep{McNutt.1996, Behr.2020}. Notably, the African wild dog is the
sole extant representative of its genus and thus considered as high-priority
species for conservation efforts \citep{Leigh.2012}. However, fragmentation and
destruction of their historic habitat may have decreased dispersal success for
this species and thus jeopardizes connectivity \citep{Leigh.2012}. Since the
impacts of climate change are not uniformly distributed across the planet, but
particularly pronounced at high latitudes and arid ecosystems, the species may
be seen as most vulnerable to such ecosystem shifts.

@@

Here, we capitalize on the previously parameterized dispersal model by
\cite{Hofmann.2022} and use dispersal simulations to investigate dispersal
success and connectivity patterns for African wild dogs under two extreme
scenarios; one assuming maximum flooding of the Okavango delta and one assuming
minimum flooding of the Okavango delta. Given that water poses a barrier to
dispersing individuals, we anticipated that dispersal prospects and connectivity
during maximum flood are low. During minimum flood, in contrast, we expect to
reveal the presence of several dispersal corridors that may be used to freely
and quickly move between remaining habitat patches.

\section{Materials and Methods}
We conducted all analyses using the programming language \textsf{R}
\citep{RCoreTeam.2022}. Any spatial data manipulation was completed using the
\textsf{terra} \citep{Hijmans.2022} and \textsf{spatstat} packages
\citep{Baddeley.2015}. Several helper functions for the simulation algorithm
were written in \textsf{C++} and imported to R using the \textsf{Rcpp} package
\citep{Eddelbuettel.2011}. Figures were generated using \textsf{ggplot2}
\citep{Wickham.2016}, \textsf{igraph} \citep{Csardi.2006}, and
\textsf{ggnetwork} \citep{Briatte.2021}. All R-scripts used to conduct our
analyses are provided through an online repository.

@@

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_Distance.png}
  \caption{Caption}
  \label{Distance}
  \end{center}
\end{figure}
